---
title: 'A Statistical Analysis of Climate in Minnesota'
author: "Caitlin Davies"
date: "1/20/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library("dplyr")
library("tidyr")
library("ggplot2")
library("devtools")
devtools::install_github("twitter/AnomalyDetection")
library("AnomalyDetection")
library("cluster")
library("scatterplot3d")
library("ggrepel")
library("knitr")
library("kableExtra")
```

```{r functions}

## This function allows me to convert month/day to the # day of the year, so that data across years can be easily compared to each other.

day_in_year <- function(m, d) {ifelse(m == 1, d,
    ifelse(m == 2, 31 + d,
    ifelse(m == 3, sum(31, 28) + d,
    ifelse(m == 4, sum(31, 28, 31) + d,
    ifelse(m == 5, sum(31, 28, 31, 30) + d,
    ifelse(m == 6, sum(31, 28, 31, 30, 31) + d,
    ifelse(m == 7, sum(31, 28, 31, 30, 31, 30) + d,
    ifelse(m == 8, sum(31, 28, 31, 30, 31, 30, 31) + d,
    ifelse(m == 9, sum(31, 28, 31, 30, 31, 30, 31, 31) + d,
    ifelse(m == 10, sum(31, 28, 31, 30, 31, 30, 31, 31, 30) + d,
    ifelse(m == 11, sum(31, 28, 31, 30, 31, 30, 31, 31, 30, 31) + d, 
           sum(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30) + d)))))))))))}

## This function allows me to group together winter tempuratures, so that data can be analyzed and visualized more easily.

group_cold <- function(t) {ifelse(t > 300, (t - 365), t)}
```

```{r import}

Stations <- read.csv(file = "MRCC Station Details.csv")
ADA <- read.csv("ADA_clean.csv")
AGASSIZ_REFUGE <- read.csv("AGASSIZ_REFUGE_clean.csv")
ARGYLE <- read.csv("ARGYLE_clean.csv")
BEMIDJI_MUNICIPAL_AP <- read.csv("BEMIDJI_MUNICIPAL_AP_clean.csv")
LEECH_LAKE <- read.csv("LEECH_LAKE_clean.csv")
POKEGAMA_DAM <- read.csv("POKEGAMA_DAM_clean.csv")
GRAND_MARAIS <- read.csv("GRAND_MARAIS_clean.csv")
DULUTH <- read.csv("DULUTH_clean.csv")
TWO_HARBORS <- read.csv("TWO_HARBORS_clean.csv")
ARTICHOKE_LAKE_1_E <- read.csv("ARTICHOKE_LAKE_1_E_clean.csv")
BEARDSLEY <- read.csv("BEARDSLEY_clean.csv")
AITKIN_2E <- read.csv("AITKIN_2E_clean.csv")
ALBERT_LEA_3_SE <- read.csv("ALBERT_LEA_3_SE_clean.csv")
AMBOY <- read.csv("AMBOY_clean.csv")

```

## Climate Stations : Initial Analysis and Visualization

----

``` {r station_observations, message = FALSE}

Stations.Observations.Count <- c(nrow(ADA), 
                                 nrow(AGASSIZ_REFUGE), 
                                 nrow(ARGYLE), 
                                 nrow(BEMIDJI_MUNICIPAL_AP), 
                                 nrow(LEECH_LAKE), 
                                 nrow(POKEGAMA_DAM), 
                                 nrow(GRAND_MARAIS), 
                                 nrow(DULUTH), 
                                 nrow(TWO_HARBORS), 
                                 nrow(ARTICHOKE_LAKE_1_E), 
                                 nrow(BEARDSLEY), 
                                 nrow(AITKIN_2E), 
                                 nrow(ALBERT_LEA_3_SE), 
                                 nrow(AMBOY))

Stations.Observations.Count.DF <- data.frame("Station.Name" = Stations$Station.Name, "Observations" =  Stations.Observations.Count)

Stations <- inner_join(Stations, Stations.Observations.Count.DF)

```

The National Centers for Environmental Information ("NOAA") divides Minnesota into 9 [Climate Divisions](https://www.ncdc.noaa.gov/monitoring-references/maps/us-climate-divisions.php), starting with Climate Division 1 in the northwest, and ending with Climate Division 9 in the southeast.


Using the [Import and Clean MRCC Datasets](https://github.com/sanity4sale/Springboard/blob/master/Import%20and%20Clean%20MRCC%20Datasets.R) R code, I have collected datasets from `r nrow(Stations)` Climate Stations throughout Minnesota, listed below.

Raw MRCC datasets contain some observations with no data, which are removed during the importing and cleaning process.

### List of Climate Stations

``` {r station_table}

Stations %>%
  select(Station.Name, County, Climate.Division, Latitude, Longitude, Elevation, Observations) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Visualizing Climate Stations

Climate Stations collected for this report can be visualized by Latitude and Longitude (with Climate Division represented by color):

```{r stations_graph1}

ggplot(Stations, aes(x = Longitude, y = Latitude, label = Station.Name)) +
  geom_point(size = 3) +
  geom_label_repel(fill = Stations$Climate.Division, segment.color = '#000000', fontface = 'bold', color = 'white')

```

And, with Elevation, the Climate Stations can be visualized in 3D:

```{r}
with(Stations, {s3d <- scatterplot3d(Longitude, Latitude, Elevation,
                 color=Stations$Climate.Division, pch=19,
                 type="h",
                 xlab="Longitude",
                 ylab="Latitude",
                 zlab="Elevation")
    s3d.coords <- s3d$xyz.convert(Longitude, Latitude, Elevation)
    text(s3d.coords$x, s3d.coords$y,
         labels= Stations$Station.Name,
         cex=.75, pos = 4)
})
```

With these graphs, we can see that Climate Division only accounts for Latitude and Longitude, and that there can be variability in Elevation within a Climate Division. This 3rd dimension might be an important factor when observing climate trend, and therefore Climate Division should not be the only grouping applied to the data. 

### Visualizing the Datasets

The cleaned datasets, on average, contain `r format(mean(Stations$Observations), big.mark=",")` observations - but the average hides the wide variability in dataset size across Climate Division, as illustrated by this box and whiskers graph:

```{r stations_boxplot}

ggplot(Stations, aes(x = factor(Climate.Division), y = Observations)) +
  geom_boxplot(fill = "#4271AE", alpha = 0.7) +
  labs(x = "Climate Divisions", title = "Dataset Size, by Climate Division")

```

Moreover, the cleaned datasets comprise different timeframes:

```{r station_years, message = FALSE}

Stations.Observations.Year.Min <- c(min(ADA$Year), 
                                 min(AGASSIZ_REFUGE$Year), 
                                 min(ARGYLE$Year), 
                                 min(BEMIDJI_MUNICIPAL_AP$Year), 
                                 min(LEECH_LAKE$Year), 
                                 min(POKEGAMA_DAM$Year), 
                                 min(GRAND_MARAIS$Year), 
                                 min(DULUTH$Year), 
                                 min(TWO_HARBORS$Year), 
                                 min(ARTICHOKE_LAKE_1_E$Year), 
                                 min(BEARDSLEY$Year), 
                                 min(AITKIN_2E$Year), 
                                 min(ALBERT_LEA_3_SE$Year), 
                                 min(AMBOY$Year))

Stations.Observations.Year.Max <- c(max(ADA$Year), 
                                 max(AGASSIZ_REFUGE$Year), 
                                 max(ARGYLE$Year), 
                                 max(BEMIDJI_MUNICIPAL_AP$Year), 
                                 max(LEECH_LAKE$Year), 
                                 max(POKEGAMA_DAM$Year), 
                                 max(GRAND_MARAIS$Year), 
                                 max(DULUTH$Year), 
                                 max(TWO_HARBORS$Year), 
                                 max(ARTICHOKE_LAKE_1_E$Year), 
                                 max(BEARDSLEY$Year), 
                                 max(AITKIN_2E$Year), 
                                 max(ALBERT_LEA_3_SE$Year), 
                                 max(AMBOY$Year))

Stations.Observations.Min.Max.DF <- data.frame("Station.Name" = Stations$Station.Name, "Min.Year" =  Stations.Observations.Year.Min, "Max.Year" = Stations.Observations.Year.Max)

Stations <- inner_join(Stations, Stations.Observations.Min.Max.DF)

ggplot(Stations, aes(Station.Name)) +
  geom_point(aes(y = Stations$Min.Year, color = "blue")) +
  geom_point(aes(y = Stations$Max.Year, color = "red")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "", y = "", title = "Date Range of Collected Datasets") +
  scale_color_manual(labels=c("First Year in Dataset", "Last Year in Dataset"), values = c("blue", "red")) +
  theme(legend.position="top", legend.title=element_blank())

```

```{r station.clusters, eval = FALSE}

Stations.cluster <- Stations[, c(4:6)]
Stations.cluster <- scale(Stations.cluster)

wssplot <- function(data, nc=6, seed=1234){
	              wss <- (nrow(data)-1)*sum(apply(data,2,var))
               	      for (i in 2:nc){
		        set.seed(seed)
	                wss[i] <- sum(kmeans(data, centers=i)$withinss)}
	                
		      plot(1:nc, wss, type="b", xlab="Number of Clusters",
	                        ylab="Within groups sum of squares")
	   }

wssplot(Stations.cluster)

k = 6
set.seed(1)
fit.km <- kmeans(Stations.cluster, centers = k, iter.max = 1000)

Stations.Results <- table(fit.km$cluster, Stations$Climate.Division)
View(Stations.Results)

clusplot(Stations.Results, clus = Stations.Results[2, ])

```

```{r ADA}
ADA.complete.years <- ADA %>% 
  select(Year, Month, Day) %>%
  group_by(Year) %>%
  summarise(
    First.Month = first(Month), 
    First.Day = first(Day),
    Last.Month = last(Month), 
    Last.Day = last(Day)
  ) %>%
  filter(First.Month == 1 & First.Day == 1 & Last.Month == 12 & Last.Day == 31)



ADA.max <- ADA %>% 
  group_by(Year) %>%
  select(Year, Month, Day, TMAX) %>%
  filter(min_rank(desc(TMAX)) <= 1) %>%
  filter((Year %in% ADA.complete.years$Year))

ADA.max$Day_of_Year <- day_in_year(ADA.max$Month, ADA.max$Day)

ADA.max.chart <- ggplot(ADA.max, aes(x = Year, y = Day_of_Year)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Hottest Day of the Year through Time", y = "Day of the Year") +
  theme(plot.title = element_text(hjust = 0.5))



ADA.min <- ADA %>% 
  group_by(Year) %>%
  select(Year, Month, Day, TMIN) %>%
  filter(min_rank(TMIN) <= 1) %>%
  filter((Year %in% ADA.complete.years$Year))

ADA.min$Day_of_Year <- day_in_year(ADA.min$Month, ADA.min$Day)

ADA.min$Day_of_Year <- group_cold(ADA.min$Day_of_Year)

ADA.min.chart <- ggplot(ADA.min, aes(x = Year, y = Day_of_Year)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Coldest Day of the Year through Time", y = "Day of the Year") +
  theme(plot.title = element_text(hjust = 0.5))

ADA.min_max <- as.data.frame(rbind(ADA.max, ADA.min))

ADA.min_max <- ADA.min_max %>%
  gather(Extreme, Temp, c(TMAX, TMIN))

ADA.min_max <- ADA.min_max[!is.na(ADA.min_max$Temp), ]

ADA.min_max$Day_of_Year <- group_cold(ADA.min_max$Day_of_Year)

ADA.min_max.chart1 <- ggplot(ADA.min_max, aes(Year, Day_of_Year, color = Extreme)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Coldest and Hottest Days of the Year through Time", y = "Day of the Year") +
  theme(plot.title = element_text(hjust = 0.5))

ADA.min_max.chart2 <- ggplot(ADA.min_max, aes(Year, Temp, color = Extreme)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Coldest and Hottest Temps of the Year through Time", y = "Tempurature") +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r eval = FALSE}

## This code isn't doing anything, perhaps because we don't have that many anomolies? 

ADA.max.anomaly <- ADA[, c(1, 9)]

ADA.max.anomaly <- ADA.max.anomaly[complete.cases(ADA.max.anomaly), ]

ADA.max.anomaly1 <- AnomalyDetectionTs(ADA.max.anomaly, max_anoms = 0.02, direction = "both", plot = TRUE)
ADA.max.anomaly1$plot

```

```{r ARGYLE}

ARGYLE.complete.years <- ARGYLE %>% 
  select(Year, Month, Day) %>%
  group_by(Year) %>%
  summarise(
    First.Month = first(Month), 
    First.Day = first(Day),
    Last.Month = last(Month), 
    Last.Day = last(Day)
  ) %>%
  filter(First.Month == 1 & First.Day == 1 & Last.Month == 12 & Last.Day == 31)



ARGYLE.max <- ARGYLE %>% 
  group_by(Year) %>%
  select(Year, Month, Day, TMAX) %>%
  filter(min_rank(desc(TMAX)) <= 1) %>%
  filter((Year %in% ARGYLE.complete.years$Year))

ARGYLE.max$Day_of_Year <- day_in_year(ARGYLE.max$Month, ARGYLE.max$Day)

ARGYLE.max.chart <- ggplot(ARGYLE.max, aes(x = Year, y = Day_of_Year)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Hottest Day of the Year through Time", y = "Day of the Year") +
  theme(plot.title = element_text(hjust = 0.5))

ARGYLE.min <- ARGYLE %>% 
  group_by(Year) %>%
  select(Year, Month, Day, TMIN) %>%
  filter(min_rank(TMIN) <= 1) %>%
  filter((Year %in% ARGYLE.complete.years$Year))

ARGYLE.min$Day_of_Year <- day_in_year(ARGYLE.min$Month, ARGYLE.min$Day)

ARGYLE.min$Day_of_Year <- group_cold(ARGYLE.min$Day_of_Year)

ARGYLE.min.chart <- ggplot(ARGYLE.min, aes(x = Year, y = Day_of_Year)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Coldest Day of the Year through Time", y = "Day of the Year") +
  theme(plot.title = element_text(hjust = 0.5))

ARGYLE.min_max <- as.data.frame(rbind(ARGYLE.max, ARGYLE.min))

ARGYLE.min_max <- ARGYLE.min_max %>%
  gather(Extreme, Temp, c(TMAX, TMIN))

ARGYLE.min_max <- ARGYLE.min_max[!is.na(ARGYLE.min_max$Temp), ]

ARGYLE.min_max$Day_of_Year <- group_cold(ARGYLE.min_max$Day_of_Year)

ARGYLE.min_max.chart1 <- ggplot(ARGYLE.min_max, aes(Year, Day_of_Year, color = Extreme)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Coldest and Hottest Days of the Year through Time", y = "Day of the Year") +
  theme(plot.title = element_text(hjust = 0.5))

ARGYLE.min_max.chart2 <- ggplot(ARGYLE.min_max, aes(Year, Temp, color = Extreme)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Coldest and Hottest Temps of the Year through Time", y = "Tempurature") +
  theme(plot.title = element_text(hjust = 0.5))

```

# Notes for next work session: 

Short.Years are the years where 

1) I've cut out NA data while cleaning the ADA dataset, and/or 
2) The data was missing in the first place. 

These years of partial data would skew the understanding of min/max/trend, so they should be isolated and removed. 

It looks like the hottest day is trending earlier, and the coldest day is trending later. 

```{r charts}

print(ADA.min.chart)

print(ARGYLE.min.chart)



print(ADA.max.chart)

print(ARGYLE.max.chart)



print(ADA.min_max.chart1)

print(ARGYLE.min_max.chart1)



print(ADA.min_max.chart2)

print(ARGYLE.min_max.chart2)

```










There are other datasets that I should work with in the future, once I have a solid grasp of the 1st dataset; 

```{r eval = FALSE}
AGASSIZ_REFUGE %>% 
  summarise(
    TMAX.mean = mean(TMAX, na.rm = TRUE),
    TMAX.median = median(TMAX, na.rm = TRUE),
    TMIN.mean = mean(TMIN, na.rm = TRUE),
    TMIN.median = median(TMIN, na.rm = TRUE)
  )

ARGYLE %>% 
  summarise(
    TMAX.mean = mean(TMAX, na.rm = TRUE),
    TMAX.median = median(TMAX, na.rm = TRUE),
    TMIN.mean = mean(TMIN, na.rm = TRUE),
    TMIN.median = median(TMIN, na.rm = TRUE)
  )

ARGYLE_model <- lm(TMIN ~ TMAX, data = ARGYLE)

summary(ARGYLE_model)

confint(ARGYLE_model)

```
